%% 2006-11-10 VZ fix of broad parameter: vertical->horizontal
\startcomponent ma-cb-fr-itemizations

\product ma-cb-fr

%\chapter[itemize]{Itemize}
\chapter[listes]{Listes}

% \index{itemize}
% \index{columns in itemize}
\index{listes}
\index{colonnes dans une liste}

\Command{\tex{startitemize}}
\Command{\tex{setupitemize}}
\Command{\tex{definesymbol}}
\Command{\tex{item}}
\Command{\tex{head}}

% One way of structuring your information is by way of
% enumeration or summing up items. The itemize command looks
% like this:
Les énumérations ou les listes permettent de structurer l'information
que vous souhaitez transmettre à vos lecteurs. La commande qui permet de
bâtir de telles listes ressemble à ceci :

\shortsetup{startitemize}

% For example:
Par exemple :

% \startbuffer
% \startitemize[R,packed,broad]
% \item Hasselt was founded in the 14th century.
% \item Hasselt is known as a so called Hanze town.
% \item Hasselt's name stems from a tree.
% \stopitemize
% \stopbuffer

\startbuffer
\startitemize[R,packed,broad]
% \item Hasselt was founded in the 14th century.
% \item Hasselt is known as a so called Hanze town.
% \item Hasselt's name stems from a tree.
\stopitemize
\stopbuffer


\typebuffer


% Within the \type{\startitemize} $\cdots$ \type{\stopitemize}
% pair you start a new item with \type{\item}. The space after
% \type{\item} is required. In the example above \type{R}
% specifies Roman numbering and \type{packed} keeps line
% spacing to a minimum. The parameter \type{broad} takes care
% %%of vertical spacing after the item separator. The example % VZ 2006-11-10
% of horizontal spacing after the item separator. The example
% would produce:

Pour saisir une nouvel \type{\item}, on doit se placer entre
\type{\startitemize} $\cdots$ \type{\stopitemize}. L'espace après
\type{\item} est obligatoire. Dans l'exemple ci-dessus, \type{R} indique
une numérotation en chiffres romains, en lettres capitales, tandis que
\type{packed} indique que l'espace entre les items sera réduit au
minimum possible.

\getbuffer

% Itemization is a two pass job. This means that you have to
% run your file twice for an optimal layout. The bracket pair
% contains information on item seperators and local set up
% variables.

Pour obtenir une présentation correcte des listes, deux compilations
successives sont nécessaires. La paire de crochets contient de
l'information concernant les séparateurs entre les items et la
configuration de variables locales.

% \placetable
%   [here]
%   [tab:itemsetup]
%   {Item separators in itemize.}
% \starttable[|l|l|]
% \HL
% \NC \bf Argument \NC \bf Item separator   \NC\SR
% \HL
% \NC 1            \NC $\bullet$            \NC\FR
% \NC 2            \NC $-$                  \NC\MR
% \NC 3            \NC $\star$              \NC\MR
% \NC $\vdots$     \NC $\vdots$             \NC\MR
% \NC n            \NC 1 2 3 4 $\cdots$     \NC\MR
% \NC a            \NC a b c d $\cdots$     \NC\MR
% \NC A            \NC A B C D $\cdots$     \NC\MR
% \NC r            \NC i ii iii iv $\cdots$ \NC\MR
% \NC R            \NC I II III IV $\cdots$ \NC\LR
% \HL
% \stoptable
\placetable
  [here]
  [tab:itemsetup]
  {Séparateurs d'items dans une liste.}
\starttable[|l|l|]
\HL
\NC \bf Argument \NC \bf Séparateur d'item\NC\SR
\HL
\NC 1            \NC $\bullet$            \NC\FR
\NC 2            \NC $-$                  \NC\MR
\NC 3            \NC $\star$              \NC\MR
\NC $\vdots$     \NC $\vdots$             \NC\MR
\NC n            \NC 1 2 3 4 $\cdots$     \NC\MR
\NC a            \NC a b c d $\cdots$     \NC\MR
\NC A            \NC A B C D $\cdots$     \NC\MR
\NC r            \NC i ii iii iv $\cdots$ \NC\MR
\NC R            \NC I II III IV $\cdots$ \NC\LR
\HL
\stoptable

% You can also define your own item separator by means of
% \type{\definesymbol}. For example if you try this:

Vous pouvez également définir votre propre séparateur d'item avec la
commande \type{\definesymbol}. Par exemple, si vous essayer ceci :

% \startbuffer
% \definesymbol[5][$\clubsuit$]

% \startitemize[5,packed]
% \item Hasselt was built on a riverdune.
% \item Hasselt lies at the crossing of two rivers.
% \stopitemize
% \stopbuffer

\startbuffer
\definesymbol[5][$\clubsuit$]

\startitemize[5,packed]
\item Hasselt was built on a riverdune.
\item Hasselt lies at the crossing of two rivers.
\stopitemize
\stopbuffer

\typebuffer

% You will get:
Vous obtiendrez :

\getbuffer

% Sometimes you would like to have headers within an
% enumeration. In that case you should use \type{\head} instead
% of \type{\item}.

Il est parfois souhaitable de pouvoir insérer un titre de section au
sein d'une énumération. Dans ce cas, vous devriez utiliser la commande
\type{\head} à la place de \type{\item}.

\startbuffer
Hasselt lies in the province of Overijssel and there are a number of
customs that are typical of this province.

\startitemize

\head kraamschudden \hfill (child welcoming)

      When a child is born the neighbours come to visit the new
      parents. The women come to admire the baby and the men come to
      judge the baby (if it is a boy) on other aspects.
      The neighbours will bring a {\em krentenwegge} along. A
      krentenwegge is a loaf of currant bread of about 1 \Meter\
      long. Of course the birth is celebrated with {\em jenever}.

\head nabuurschap (naberschop) \hfill (neighbourship)

      Smaller communities used to be very dependent on the
      cooperation among the members for their well being. Members of
      the {\em nabuurschap} helped each other in difficult times
      during harvest times, funerals or any hardship that fell upon
      the community.

\head Abraham \& Sarah  \hfill (identical)

      When people turn 50 in Hasselt it is said that they see Abraham
      or Sarah. The custom is to give these people a {\em speculaas}
      Abraham or a Sarah. Speculaas is a kind of hard spiced biscuit.

\stopitemize
\stopbuffer

\typebuffer

% The \type{\head} can be set up with \type{\setupitemize}.
% In case of a page breaking a \type{\head} will appear on a
% new page. (The \type {\Meter} command is explained in \in
% {chapter} [units].)

La commande \type{\head} peut être configurée avec
\type{\setupitemize}. Si une coupure de page intervient, le contenu de
la commande \type{\head} apparaîtra sur une nouvelle page. (La commande
\type{\Meter} est définie au \in{chapitre}[units].)

% The example of old customs will look like this:

L'exemple de liste des coutumes anciennes de Hasselt aura donc cet
aspect :

\getbuffer

% The set up parameters of itemize are described in
% \in{table}[tab:tablesetup].

Les paramètres de configuration des listes sont décrits dans le
\in{tableau}[tab:tablesetup].

% \placetable
%   [here,force]
%   [tab:tablesetup]
%   {Set up parameters in itemize.}
% \starttable[|l|l|]
% \HL
% \NC \bf Set up \NC \bf Meaning                                      \NC\SR
% \HL
% \NC standard   \NC standard (global) set up                  \NC\FR
% \NC packed     \NC no vertical spacing between items                \NC\MR
% \NC serried    \NC no horizontal spacing between separator and text \NC\MR
% \NC joinedup   \NC no vertical spacing before and after itemize     \NC\MR
% \NC broad      \NC horizontal spacing between separator and text    \NC\MR
% \NC inmargin   \NC place separator in margin                        \NC\MR
% \NC atmargin   \NC place separator on margin                        \NC\MR
% \NC stopper    \NC place full stop after separator                  \NC\MR
% \NC columns    \NC put items in columns                             \NC\MR
% \NC intro      \NC prevent page breaking after introduction line      \NC\MR
% \NC continue   \NC continue numbering or lettering                  \NC\LR
% \HL
% \stoptable

\placetable
  [here,force]
  [tab:tablesetup]
  {Paramètres de configuration des listes.}
\starttable[|l|l|]
\HL
\NC \bf Paramètre \NC \bf Signification                              \NC\SR
\HL
\NC standard   \NC configuration standard (globale)                  \NC\FR
\NC packed     \NC pas d'espace vertical entre les items             \NC\MR
\NC serried    \NC pas d'espace horizontal entre le séparateur et le texte \NC\MR
\NC joinedup   \NC pas d'espace vertical avant et après la liste     \NC\MR
\NC broad      \NC espace horizontal entre le séparateur et le texte \NC\MR
\NC inmargin   \NC place le séparateur dans la marge                 \NC\MR
\NC atmargin   \NC place le séparateur sur la marge                  \NC\MR
\NC stopper    \NC place un point final après le séparateur          \NC\MR
\NC columns    \NC compose les items en colonne                      \NC\MR
\NC intro      \NC empêche les coupures de page après les lignes d'introduction \NC\MR
\NC continue   \NC continue la numérotation                          \NC\LR
\HL
\stoptable

% You can use the set up parameters in \type{\startitemize},
% but for reasons of consistency you can make them valid for
% the complete document with \type{\setupitemize}.

On peut définir les paramètres de configuration dans
\type{\startitemize} mais, pour des raisons de cohérence au sein de
votre document, vous devriez utiliser la commande \type{\setupitemize},
qui définira ces paramètre de manière globale.

% The parameter \type{columns} is used in conjunction with
% a number. If you type this:

Le paramètre \type{columns} est utilisé conjointement avec un nombre. Si
vous saisissez ceci :

\startbuffer
\startitemize[n,columns,four]
\item Achter 't Werk
.
.
.
\item Justitiebastion
\stopitemize
\stopbuffer

\typebuffer

% You will get:
Vous obtiendrez

\startbuffer
\startitemize[n,columns,four]
\item Achter 't Werk
\item Baangracht
\item Brouwersgracht
\item Eikenlaan
\item Eiland
\item Gasthuisstraat
\item Heerengracht
\item Hofstraat
\item Hoogstraat
\item Julianakade
\item Justitiebastion
\stopitemize
\stopbuffer

\getbuffer

% Sometimes you want to continue the enumeration after a short
% intermezzo. Then you type for example
% \type{\startitemize[continue,columns,three,broad]} and
% numbering will continue in a three column format.

Il arrive parfois qu'on souhaite poursuivre une énumération après un
court paragaphe de texte. Il suffit, pour ce faire, de saisir (par
exemple) \type{\startitemize[continue,columns,three,broad]}, et la
numérotation des items reprendra là où elle s'était arrêtée.

\startbuffer
\startitemize[continue, columns, three, broad]
\item Kaai
\item Kalverstraat
\item Kastanjelaan
\item Keppelstraat
\item Markt
\item Meestersteeg
\item Prinsengracht
\item Raamstraat
\item Ridderstraat
\item Rosmolenstraat
\item Royenplein
\item Van Nahuijsweg
\item Vicariehof
\item Vissteeg
\item Watersteeg
\item Wilhelminalaan
\item Ziekenhuisstraat
\stopitemize
\stopbuffer

\getbuffer

% The parameter \type{broad} enlarges the horizontal space
% between item separator and itemtext.

Le paramètre \type{broad} augmente l'espace horizontal entre le
séparateur d'item et le texte de l'item.

\shortsetup{setupitemize}

% An itemize within an itemize is automatically typeset in a
% correct way. For example if you type:

Une liste dans une liste est automatiquement composée correctement. Par
exemple, si vous saisissez :

\startbuffer
In the Netherlands the cities can determine the size of a number of
taxes. So the cost of living can differ from town to town. There are
differences of up to 50\% in taxes such as:

\setupitemize[2][width=5em]
\startitemize[n]

\item real estate tax

      The real estate tax is divided into two components:

      \startitemize[a,packed]
      \item the ownership tax
      \item the tenant tax
      \stopitemize

      If the real estate has no tenant the owner pays both components.

\item dog licence fee

      The owner of one or more dogs pays a fee. When a dog has died
      or been sold the owner has to inform city hall.

\stopitemize
\stopbuffer

\typebuffer

% then the horizontal space between item separator and text at
% the second level is set with
% \type{\setupitemize[2][width=5em]}.

l'espace horizontal entre le séparateur d'item et le texte de la liste
de second niveau sera paramétré avec \type{\setupitemize[2]{width=5em}}.

% The example will look like this:

L'exemple ressemblera à cela :

\start
\getbuffer
\stop

\stopcomponent