% Translation:
\startcomponent ma-cb-cz-margintexts
\project ma-cb
\environment ma-cb-env-cz

\product ma-cb-cz

\chapter{Margin texts}

\index{margin text}

\Command{\tex{inmargin}}
\Command{\tex{inleft}}
\Command{\tex{inright}}
\Command{\tex{margintitle}}

It is very easy to put text in the margin. You just use
\type{\inmargin}.

\shortsetup{inmargin}

You may remember one of the earlier examples:

\typebuffer[marginpicture]

This would result in a figure in the \pagereference
[marginpicture]\getbuffer [marginpicture]margin. You
can imagine that it looks quite nice in some documents. But
be careful. The margin is rather small so the figure could
become very marginal.

A few other examples are shown in the text below.

\startbuffer
The Ridderstraat (Street of knights) \inmargin{Street of\\Knights}
is an obvious name. In the 14th and 15th centuries, nobles and
prominent citizens lived in this street. Some of their big houses
were later turned into poorhouses \inright{poorhouse}and old
peoples homes.

Up until \inleft[low]{\tfc 1940}1940 there was a synagog in the
Ridderstraat. Some 40 Jews gathered there to celebrate their
sabbath. During the war all Jews were deported to Westerbork and
then to the extermination camps in Germany and Poland. None of
the Jewish families returned. The synagog was knocked down in
1958.
\stopbuffer

\typebuffer

The commands \type{\inmargin}, \type{\inleft} and
\type{\inright} all have the same function. In a two sided
document \type{\inmargin} puts the margin text in the correct
margin. The \type{\\} is used for line breaking. The example
above would look like this:

\getbuffer

You can set up the margin text with:

\starttyping
\setupinmargin
\stoptyping

\stopcomponent
